Left-linear rules in adhesive categories

### 2024-02-12 icalp/fusione.tex
intro (ultimo para ancora non ok)

One of the key pay-off of concurrency theory is the idea that the behaviour of 
a computational device can be modelled by abstracting its sequences of steps
%performed by the device 
via a suitable equivalence. Intuitively, the equivalence 
captures when steps are causally unrelated and thus could be executed in any 
order, and possibly in parallel. 
%
Two seminal contributions to this line of research have been
Mazurkiewicz's traces~\cite{Mazurkiewicz86} and Winskel's event structures~\cite{NPW:PNES}.
%
Concerning formalisms based on the rewriting paradigm, i.e. where the
behaviour is dictated by a set of rules that have to be instantiated with
the system at hand, concurrency often boils down to having a notion of 
independence between two consecutive steps.
%thus viewing a sequence of steps as a concurrent computation. 
Noteworthy examples are
interchange law~\cite{Mes92} in term rewriting and permutation 
equivalence~\cite{JJL80} in $\lambda$-calculi.

@@

A noteworthy advantage is that DPO rewriting can be formulated over
adhesive categories and their variants~\cite{lack2005adhesive,ehrig2006weak}. 
%
Operating within the framework of adhesive categories allows one to devise 
the foundations of a rewriting theory that can be subsequently instantiated 
to the context of interest, thus avoiding the need of repeatedly proving
similar ad-hoc results in for each specific setting. In fact, the framework is
so general that it allowed to recast in the DPO approach many additional formalisms,
ranging from toposes~\cite{johnstone2007quasitoposes} to string diagrams~\cite{bonchi2022string}.

So far, most of the theoretical results focussed on what are called \emph{linear rules}: 
intuitively, a step is required only to consume, preserve and generate entities. 
However, in some situations it is also necessary to ``merge'' parts of the state. 
This happens already in the context of string diagram mentioned above, but also 
in graphical implementations of nominal calculi where, as a result of name passing, 
the received name is identified with a local one at the
receiver~\cite{CVY:ESSPE,Gad07} or in the visual modelling of bonding in
biological/chemical processes~\cite{PUY:MBPE}. A similar mechanisms is
at the core of e-graphs (equality graphs), recently used for
rewrite-driven compiler optimisations~\cite{WNW:egg}: rather than
explicitly merging nodes corresponding to equal expression, the idea
here -- conceptually and, as we will see, also mathematically similar --
is to maintain an equivalence\todo{P: more precisely a congruence}
over the nodes.  \todo{P: la sequenza di lavori su questo \`e ampia,
  vedere se c'\`e qualcosa d'altro}

Technically, to acquire the expressiveness needed for expressing
fusions, i.e., the merge of some elements in the state, requires to
move from linear to left-linear rewriting rules. While left-linear
rules have been considered in various instances (mainly in categories
of graphs) and some results concerning the corresponding rewriting
theory have been put forward, to the best of our knowledge the
phenomena arising when devising a theory of indepedence and
concurrency for rewriting with left-linear rules in the general
setting of adhesive and quasi-adhesive categories
%(which capture most ``applications'') 
have not been systematically
explored~\cite{Ehrig1976,EHP:BRfTToHLRS,baldan2011adhesivity} \todo{P: see also \cite{HMP:DPORADD,BHK:CTNL} ... verificare e aggiungere}

Here we perform this exploration working at the level of $\mathcal{M}$-adhesive catgories, which encompass ...

As an initial step in performing this exploration, we observe that for
left-linear rules a number of fundamental properties that play a role
in the corresponding theory of rewriting need to be deeply revised, or
plainly do not hold.
%
Firstly, the notion of sequential independence and Church Rosser
theorem, ensuring that sequential independent steps can be performed
in reverse order, does not hold out of the box. Even worse, given two
independent steps it might be possible to switch them in several
different ways. As we will observe, this might prevent the development
of a sensible theory of concurrency as switches are used as a basis
witch equivalence of derivation sequences, identifying rewriting
sequences of steps which differs only for the order of independent
steps.

We single a well-behaved class of left-linear rewriting
systems referred to as \emph{well-switching rewriting systems}, where sequential independence ensures existence and uniqueness of the
switch, and argue that this is the right setting for dealing with
left-linear rewriting systems.

On the one hand, we argue that they capture most categories of
interest for rewriting. In particular, we single out a class of
left-linear rewriting systems on graph-like structures which are very
well-switchable and we show that rewriting systems on graphs with
equivalences, similar in spirit to e-graphs are also
well-switchable. As a sanity check, also linear rewriting systems on
$\mathcal{M}$-adhesive categories are well switching.

On the other hand, we show that an appropriate theory of rewriting can
be developed for well switching rewriting systems, recovering, sometimes in
weakened form, results which hold in the linear case.  Church-Rosser
is fully recovered for well-switchable rewriting systems.

There is, however, a further, serious problem.  For linear
rewriting systems s, can be characterised in terms of the
existence of what is called a consistent permutation~\cite{xxx}
(roughly, an isomorphism between their colimits). This result, which
is folklore for various categories of graphs, is explicitly proved
here for $\mathcal{M}$-adhesive categories.
%
The characterisation of switch equivalence based on consistent
permutations is the key to derive some
properties for linear rewriting systems, namely
\begin{enumerate}
\item \emph{confluence of switching}: when transforming derivation
  sequences into equivalent ones by sqwitching independent steps, the
  result not depends on the order in which switchings are applied, but
  only on the resulting permutation;

>>>>

One of the key pay-off of concurrency theory is the idea that the behaviour of 
a computational device can be modelled by abstracting its sequences of steps
%performed by the device 
via a suitable equivalence. Intuitively, the equivalence 
captures when 
%sequential 
steps are causally unrelated and thus could be executed in any 
order, and possibly in parallel. 
%
Two seminal contributions to this line of research have been
Mazurkiewicz's traces~\cite{Mazurkiewicz86} and Winskel's event structures~\cite{NPW:PNES}.
%
Concerning formalisms based on the rewriting paradigm, i.e. where the
behaviour is dictated by a set of rules that have to be instantiated with
the system at hand, concurrency often boils down to having a notion of 
independence between two consecutive steps.
%thus viewing a sequence of steps as a concurrent computation. 
Noteworthy examples are
interchange law~\cite{Mes92} in term rewriting and permutation 
equivalence~\cite{JJL80} in $\lambda$-calculi.

@@

A noteworthy advantage is that DPO rewriting can be formulated over
adhesive categories and their variants~\cite{lack2005adhesive,ehrig2006weak}. 
%
Operating within the framework of adhesive categories allows one to devise 
the foundations of a rewriting theory that can be subsequently instantiated 
to the context of interest, thus avoiding the need of repeatedly proving
similar ad-hoc results for each specific setting. In fact, the framework is
so general that it allowed to recast in the DPO approach many additional formalisms,
ranging from toposes~\cite{johnstone2007quasitoposes} to string diagrams~\cite{bonchi2022string}.

So far, most of the theoretical results focussed on what are called \emph{linear rules}: 
intuitively, a step is required only to consume, preserve and generate entities. 
However, in some situations it is also necessary to ``merge'' parts of the state. 
This happens already in the context of string diagram mentioned above, but also 
in graphical implementations of nominal calculi where, as a result of name passing, 
the received name is identified with a local one at the
receiver~\cite{CVY:ESSPE,Gad07} or in the visual modelling of bonding in
biological/chemical processes~\cite{PUY:MBPE}. A similar mechanisms is
at the core of e-graphs (equality graphs), recently used for
rewrite-driven compiler optimisations~\cite{WNW:egg}: rather than
explicitly merging nodes corresponding to equal expression, the idea
here -- conceptually and, as we will see, also mathematically similar --
is to maintain an equivalence %\todo{P: more precisely a congruence}
over the nodes.  
%\todo{P: la sequenza di lavori su questo \`e ampia,
%vedere se c'\`e qualcosa d'altro}

Technically, to acquire the expressiveness needed for capturing
fusions, i.e. the merge of some elements in the state, requires to
move from linear to left-linear rewriting rules. While left-linear
rules have been considered in various instances (mainly in categories
of graphs) and some results concerning the corresponding rewriting
theory have been put forward, to the best of our knowledge the
phenomena arising when devising a theory of indepedence and
concurrency for rewriting with left-linear rules in the general
setting of adhesive and quasi-adhesive categories
%(which capture most ``applications'') 
have not been systematically
explored~\cite{Ehrig1976,EHP:BRfTToHLRS,baldan2011adhesivity} \todo{P: see also \cite{HMP:DPORADD,BHK:CTNL} ... verificare e aggiungere}

Here we perform this exploration working at the level of $\mathcal{M}$-adhesive categories, 
%i.e.such that adhesivity holds only for a class $\mathcal{M}$ of arrows, 
which encompass a large family of DPO-based formalisms adopted by pratictioners~\cite{xxx}.

As an initial step, % in performing this exploration, 
we observe that for
left-linear rules a number of fundamental properties that play a role
in the concurrency theory of linear rewriting need to be deeply revised, or
plainly do not hold.
%
Firstly, the notion of sequential independence and Church Rosser
theorem, ensuring that sequential independent steps can be performed
in reverse order, does not hold out of the box. Even worse, given two
independent steps it might be possible to switch them in several
different ways. As we will observe, this might prevent the development
of a sensible theory of concurrency as switches are used as a basis
for equating sequences of rewriting steps that differs only for the order of independent
steps.

We single out a class of left-linear rewriting
systems, referred to as \emph{well-switching rewriting systems}, where sequential independence ensures existence and uniqueness of the
switch, and argue that this is the right setting for dealing with
left-linear rewriting systems.

On the one hand, they capture most categories of
interest for rewriting. In particular, we single out a class of
left-linear rewriting systems on graph-like structures that are
well-switching and we show that rewriting systems on graphs with
equivalences are so. 
As a sanity check, we prove that linear rewriting systems on
$\mathcal{M}$-adhesive categories are well-switching.

On the other hand, we show that a concurrency theory of rewriting can
be developed for well-switching rewriting systems, recovering, sometimes in
weakened form, results which hold in the linear case.  Specifically, 
the Church-Rosser theorem is fully recovered.
%for well-switchable rewriting systems.

There is, however, a further problem.  For linear
rewriting systems, switch equivalence can be characterised in terms of the
existence of what is called a consistent permutation~\cite{xxx}.
%(roughly, an isomorphism between their colimits). 
This result, which
is folklore for various categories of graphs, is explicitly proved
here for $\mathcal{M}$-adhesive categories.
%
The characterisation of switch equivalence based on consistent
permutations is the key to derive some
properties for linear rewriting systems, namely
\begin{enumerate}
\item \emph{confluence of switching}: when transforming
  sequences of rewriting steps into equivalent ones by switching independent steps, the
  result not depends on the order in which switchings are applied, but
  only on the resulting permutation;
